---
title: "A very short guide on data.frames"
author: "Even Garvang"
date: "Spring 2020"
output: 
  bookdown::html_document2:
    toc: true
    toc_float: true
    highlight: pygments
editor_options: 
  chunk_output_type: console
---

<style type="text/css">

.subtitle { font-size: 24px; }

.author { font-size: 16px; 
  font-style: italic;
  }

.date { font-size: 16px; 
  font-style: italic;
  }

h1 { font-size: 28px; }

h2 { font-size: 24px; }

h3 { font-size: 20px; }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

A data.frame is arguably the most important kind of object you work with in R as a biologist. This very short tutorial covers some important aspects of handling data.frames, namely adding and extracting information, and subsetting.

We will make this example data.frame for working with throughout:

```{r}
my_df <- data.frame(
  numbers_1 = 1:5,
  numbers_2 = 6:10,
  letters_1 = letters[1:5],
  letters_2 = letters[6:10],
  stringsAsFactors = FALSE
)
my_df
```


# Accessing data within your data.frame

You can extract data in two main ways: with the `$` operator or with square brackets `[]`.

The `$` operator extracts a column from the data frame into a vector:

```{r}
my_df$numbers_1
my_df$letters_1
```

Square brackets can take either rows or columns, as well as single data entries. The syntax is `data[row,column]`. If you leave either of the fields within the brackets empty, it will return all the rows or columns, respectively.

```{r}
# extract data from row 2, column 4:
my_df[2,4]
#extract row 2, column 1 and 3:
my_df[2, c(1,3)]

# extract row 2, all columns:
my_df[2,]
# extract column 4, all rows:
my_df[,4]

```

Note that the last one is the same as writing `my_data$letters_2`.

You can also supply a column name in the brackets to extract one or more column (remember quotation marks):

```{r}
my_df["letters_1"]
my_df[c("numbers_2", "letters_2")]
```

# Changing data and adding columns
## Changing data

Luckily, the process for changing a data point is very similar to extracting one. All you need to do is to put the indexing on the left side of `<-` (or `=`).

```{r}
# change a single value
my_df[2,4] <- "changed"
my_df

# change a column
my_df$numbers_2 <- c(1,1,2,3,5)
my_df

# change a row
my_df[1,] <- c("this", "row", "is", "changed")
my_df
```

Notice how this changes the original data, so use this with caution!

```{r}
# make a new df since we broke the last one ...
my_df <- data.frame(
  numbers_1 = 1:5,
  numbers_2 = 6:10,
  letters_1 = letters[1:5],
  letters_2 = letters[6:10],
  stringsAsFactors = FALSE
)
```


## Adding columns
In the same way you can write e.g. `my_df$numbers_2 <-` to overwrite the column `numbers_2`, you can create a new column by writing a name that doesn't already exist:

```{r}
# new column
my_df$numbers_3 <- 11:15
my_df
```

A very useful application if this is that you can compute values in a new column using data from other columns:

```{r}
my_df$numbers_4 <- my_df$numbers_1 + my_df$numbers_3
my_df
```

This kind of vectorized operations is among the most useful features of R, and it is essential for handling data.

# Subsetting
Subsetting is similar to extracting data in that you select out certain rows (and columns). You can use either the square brackets or R's `subset` function for this. A difference is that when subsetting, you often want to select rows based on some defined criteria, and often you use R's logical operators, like `==`, `>` or `%in%` for this. This is based on a different way of extracting data than we covered before, namely **providing a vector of `TRUE` and `FALSE` to select all row numbers that are `TRUE`.** For example, for selecting rows 1 and 2, you could write:

```{r}
my_df[1:2,]
```

But you could also write:

```{r}
my_df[c(TRUE,TRUE,FALSE,FALSE,FALSE),]
```

Then, consider that writing:

```{r}
my_df$numbers_1 < 3
```

provides the exact same vector that we used earlier for subsetting. This means that you can write:

```{r}
my_df[my_df$numbers_1 < 3,]
```

to select all rows where the `numbers_1` column is smaller than 3.

You can select based on several criteria using the `%in%` operator, which checks if the value of your data is equal to a value in a vector you provide, e.g.:

```{r}
my_df[my_df$letters_1 %in% c("a", "b"),]
```

I won't cover the logical operators comprehensively here, but a lot of material is available online. It is as simple as: **if you are able to create your subsetting criteria as a logical expression in R, you will have no trouble subsetting.**

An alternative to using square brackets is using the `subset` function. This works in the same way, with logical operators, but a different syntax.

```{r}
subset(my_df, numbers_1 < 3)

subset(my_df, letters_1 %in% c("a", "b"))
```

If you want to know more about the `subset` function, try accessing it's help page by writing `?subset`.